---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}
description <- readLines("DESCRIPTION")
version <- gsub(" ", "", gsub("Version:", "", grep("Version:", description, value = TRUE)))
```

```{r, echo = FALSE}
description <- readLines("DESCRIPTION")
rvers <- stringr::str_match(grep("R \\(", description, value = TRUE), "[0-9]{1,4}\\.[0-9]{1,4}\\.[0-9]{1,4}")[1,1]
```

#### General
[![Build Status](https://travis-ci.org/pat-s/oddsratio.svg?branch=master)](https://travis-ci.org/pat-s/oddsratio)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![codecov](https://codecov.io/gh/pat-s/oddsratio/branch/master/graph/badge.svg)](https://codecov.io/gh/pat-s/oddsratio)
[![minimal R version](https://img.shields.io/badge/R%3E%3D-`r rvers`-6666ff.svg)](https://cran.r-project.org/)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-yellowgreen.svg)](/commits/master)

#### CRAN
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/oddsratio)](http://cran.r-project.org/package=oddsratio)
[![Downloads](http://cranlogs.r-pkg.org/badges/oddsratio?color=brightgreen)](http://www.r-pkg.org/pkg/oddsratio)
[![Rdoc](http://www.rdocumentation.org/badges/version/oddsratio)](http://www.rdocumentation.org/packages/oddsratio)

#### Github
[![Github Releases](https://img.shields.io/github/downloads/pat-s/oddsratio/latest/total.svg)](https://github.com/pat-s/oddsratio)
[![packageversion](https://img.shields.io/badge/Package%20version-`r version`-orange.svg?style=flat-square)](commits/master)


# oddsratio

Functions for calculation and plotting of odds ratios of Generalized Additive (Mixed)
Models and Generalized Linear (Mixed) Models with a binomial
response variable (i.e. logistic regression models).  

## Examples

### GLM

Odds ratio calculation of predictors `gre` & `gpa` of a fitted model `fit.glm`
with increment steps of 380 and 5, respectively.  
For factor variables (here: `rank` with 4 levels), automatically all odds ratios
corresponding to the base level (here: `rank1`) are returned including their 
respective confident intervals. The default level is 95%. 
However, other levels can be specified with the param `CI`.
Data source: http://www.ats.ucla.edu/stat/r/dae/logit.htm

```{r}
library(oddsratio)
dat <- read.csv("http://www.ats.ucla.edu/stat/data/binary.csv")
dat$rank <- factor(dat$rank)
fit.glm <- glm(admit ~ gre + gpa + rank, data = dat, family = "binomial")

calc.oddsratio.glm(data = dat, model = fit.glm, 
                   incr = list(gre = 380, gpa = 5, CI = 0.95))
```

### GAM

For GAMs, the calculation of odds ratio is different. 
Due to its non-linear definition, odds ratios do only apply to specific
value changes and are not constant throughout the whole value range of the 
predictor as for GLMs. 
Hence, odds ratios of GAMs can only be computed for one predictor at a time by 
holding all other predictors at a fixed value while changing the value of the 
specific predictor. 
Confident intervals are currently fixed to the 95% level for GAMs. 
Data source: `?mgcv::predict.gam()`

Here, the usage of `calc.oddsratio.gam()` is shown by calculating odds ratios of 
pred `x2` for a 20% steps across the whole value range of the predictor.

```{r}
suppressPackageStartupMessages(library(mgcv))
set.seed(1234)
n <- 200
sig <- 2
dat <- gamSim(1, n = n,scale = sig, verbose = FALSE)
dat$x4 <- as.factor(c(rep("A", 50), rep("B", 50), rep("C", 50), rep("D", 50)))
fit.gam <- mgcv::gam(y ~ s(x0) + s(I(x1^2)) + s(x2) + offset(x3) + x4, data = dat)

calc.oddsratio.gam(data = dat, model = fit.gam, pred = "x2", 
                   percentage = 20, slice = TRUE)
```

If you want to compute a single odds ratio for specific values, simply set 
param `slice = FALSE`:

```{r}
calc.oddsratio.gam(data = dat, model = fit.gam, 
                   pred = "x2", values = c(0.099, 0.198))
```

From version 0.3.0 on (Github version 0.2.0.9000) you can now do the following things:

* Plot smoothing functions of GAM(M)s

```{r}
plot_smooth.gam(fit.gam, pred = "x2", title = "Predictor 'x2'")
```

* Insert the calculated odds ratios into the smoothing function

```{r}
plot.object <- plot_smooth.gam(fit.gam, pred = "x2", title = "Predictor 'x2'")
or.object <- calc.oddsratio.gam(data = dat, model = fit.gam, 
                   pred = "x2", values = c(0.099, 0.198))

plot <- add.or.to.plot(plot.object, or.object, height.or = 5, x.shift = 0.04)
plot
```

If you want to insert multiple odds ratios, just go on!

```{r}
or.object2 <- calc.oddsratio.gam(data = dat, model = fit.gam, 
                   pred = "x2", values = c(0.4, 0.6))

add.oddsratio.into.plot(plot, or.object2, height.or = 2.5, x.shift = 0.024, 
               col.line = "green4", col.text = "green4")
```


## Installation

Install from CRAN:

```R
install.packages("oddsratio")
```

Get the development version from Github:

```R
devtools::install_github("pat-s/oddsratio", build_vignettes = TRUE)
```
## Examples

To see full examples, please see the examples in the respective help pages 
`?calc.oddsratio.gam` and `?calc.oddsratio.gam` or take a look at the package
vignette.
