---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[![Build Status](https://travis-ci.org/pat-s/oddsratio.svg?branch=master)](https://travis-ci.org/pat-s/oddsratio)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/oddsratio)](http://cran.r-project.org/package=oddsratio)
[![Downloads](http://cranlogs.r-pkg.org/badges/oddsratio?color=brightgreen)](http://www.r-pkg.org/pkg/oddsratio)

# oddsratio

Convenience functions for odds ratio calculation of Generalized Additive (Mixed)
Models and Generalized Linear (Mixed) Models with a binomial
response variable (i.e. logistic regression models).  

## Examples

### GLM

Odds ratio calculation of predictors `gre` & `gpa` of a fitted model `fit.glm`
with increment steps of 380 and 5, respectively.  
For factor variables (here: `rank` with 4 levels), automatically all odds ratios
corresponding to the base level (here: `rank1`) are returned.  
Data source: http://www.ats.ucla.edu/stat/r/dae/logit.htm

```{r}
library(oddsratio)
dat <- read.csv("http://www.ats.ucla.edu/stat/data/binary.csv")
dat$rank <- factor(dat$rank)
fit.glm <- glm(admit ~ gre + gpa + rank, data = dat, family = "binomial")

calc.oddsratio.glm(data = dat, model = fit.glm, incr = list(gre = 380, gpa = 5))
```

### GAM

For GAMs, the calculation of odds ratio is different. 
Due to its non-linear definition, odds ratios do only apply to specific
value changes and are not constant throughout the whole value range of the 
predictor as for GLMs. 
Hence, odds ratios of GAMs can only be computed for one predictor at a time by 
holding all other predictors at a fixed value while changing the value of the 
specific predictor. 
Data source: `?mgcv::predict.gam()`

Here, the usage of `calc.oddsratio.gam()` is shown by calculating odds ratios of 
pred `x2` for a 20% steps across the whole value range of the predictor.

```{r}
suppressPackageStartupMessages(library(mgcv))
set.seed(1234)
n <- 200
sig <- 2
dat <- gamSim(1, n = n,scale = sig, verbose = FALSE)
dat$x4 <- as.factor(c(rep("A", 50), rep("B", 50), rep("C", 50), rep("D", 50)))
fit.gam <- mgcv::gam(y ~ s(x0) + s(I(x1^2)) + s(x2) + offset(x3) + x4, data = dat)

calc.oddsratio.gam(data = dat, model = fit.gam, pred = "x2", 
                   percentage = 20, slice = TRUE)
```

If you want to compute a single odds ratio for specific values, simply set 
param `slice = FALSE`:

```{r}
calc.oddsratio.gam(data = dat, model = fit.gam, 
                   pred = "x2", values = c(0.099, 0.198))
```


## Installation

Install from CRAN:

```R
install.packages("oddsratio")
```

Get the development version from Github:

```R
devtools::install_github("pat-s/oddsratio", build_vignettes = TRUE)
```
## Examples

To see full examples, please see the examples in the respective help pages 
`?calc.oddsratio.gam` and `?calc.oddsratio.gam` or take a look at the package
vignette.

## To Do

- Add calculation of odds ratio confidence intervals
- Implement plotting function of calculated odds ratio in GLM/GAM function?
- Add references for odds ratio calculation?
