---
title: "exemplary usage of functions"
author: "Patrick Schratz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{nlme}
---

```{r, echo=FALSE}
calc.oddsratio.gam <- function(data, model, pred, values, percentage, 
                               slice = FALSE, quietly = FALSE) {
  
  names.pred <- colnames(data)
  
  if (slice) {
    
    # get range of pred distribution and slice in x tiles
    range <- max(data[, pred]) - min(data[, pred])
    step <- range / (100/percentage)
    
    range.v <- c()
    
    # create vector of percentage
    for (i in 0:(100/percentage)) {
      range.v <- c(range.v, min(data[, pred]) + step * i)
    }
    
    result <- data.frame(odds.ratio = numeric(length = 100/percentage),
                         from = numeric(length = 100/percentage),
                         to = numeric(length = 100/percentage),
                         perc_from = character(length = 100/percentage),
                         perc_to = character(length = 100/percentage),
                         stringsAsFactors = FALSE)
    
    if (!quietly)
      # print variable information
      cat("Predictor: '", pred, "'\nSteps:     ", 100/percentage,
          " (", percentage, "%)\n", sep = "")
    
    # apply OR calc for vector
    for (x in 1:(100/percentage)) {
      
      # set all preds to their mean if they are numeric
      for (i in names.pred) {
        if (is.numeric(data[[i]])) 
           data[[i]] <- mean(data[[i]])
      }
      
      # reduce to 1 row
      data <- data[1, ]
      # subset DF to preds only
      data <- data[, names.pred]
      
      # set values[1] of pred
      data[, pred] <- range.v[x]
      # calc log odds for value 1
      pred.gam1 <- as.numeric(predict(model, data, type = "link"))
      # set values[2] of pred
      data[, pred] <- range.v[x + 1]
      # calc log odds for value 2
      pred.gam2 <- as.numeric(predict(model, data, type = "link"))
      odds.ratio <- as.numeric(predict(model, data, type = "link"))
      
      # combine results in DF
      result$odds.ratio[x] <- as.numeric(exp(pred.gam2 - pred.gam1), 2)
      result$from[x] <- round(range.v[x], 3)
      result$to[x] <-  round(range.v[x + 1], 3)
      result$perc_from[x] <- as.character((percentage*x - percentage))
      result$perc_to[x] <- as.character((percentage)*x)

      if (!quietly) {
        cat("\nOdds ratio from ", round(range.v[x], 3), "(", 
            (percentage*x - percentage), "%)", " to ", 
            round(range.v[x + 1], 3), 
            "(", (percentage)*x, "%): ", 
            as.numeric(exp(pred.gam2 - pred.gam1), 2), sep = "")
      }
    }
    return(invisible(result))
  }
  
  # set all preds to their mean if they are numeric
  for (i in names.pred) {
    if (is.numeric(data[[i]]))
       data[[i]] <- mean(data[[i]])
  }
  
  # reduce to 1 row
  data <- data[1, ]
  # subset DF to preds only
  data <- data[, names.pred]
  
  # set values[1] of pred
  data[, pred] <- values[1]
  # calc log odds for value 1
  pred.gam1 <- as.numeric(predict(model, data, type = "link"))
  # set values[2] of pred
  data[, pred] <- values[2]
  # calc log odds for value 2
  pred.gam2 <- as.numeric(predict(model, data, type = "link"))
  
  odds.ratio <- as.numeric(exp(pred.gam2 - pred.gam1), 2)
  value.info <- list(paste0("OR (", values[1], " to ", values[2], ")"))
  
  if (!quietly & !slice)
    # print variable information
    cat("Predictor: '", pred, "'\n", sep = "")
  
  if (!quietly & !slice) {
    cat("\nOdds ratio from '", values[1], "' to '",
        values[2], "': ", 
        as.numeric(exp(pred.gam2 - pred.gam1), 2), sep = "" ) 
  }
}

```


# Load example data

Source: help() of predict.gam()

```{r}
suppressPackageStartupMessages(library(mgcv))
n <- 200
sig <- 2
dat <- suppressMessages(gamSim(1, n = n,scale = sig))
dat$x4 <- as.factor(c(rep("A", 50), rep("B", 50), rep("C", 50), rep("D", 50)))

fit_gam <- gam(y ~ s(x0) + s(I(x1^2)) + s(x2) + offset(x3) + x4, data = dat)
```

# Examples {.tabset}

## Calculate OR for specific increment step of continuous variable

```{r}
calc.oddsratio.gam(model = fit_gam, data = dat, pred = "x2",
                   values = c(0.099, 0.198))
```

## Calculate OR for change of indicator variable

```{r}
calc.oddsratio.gam(model = fit_gam, data = dat, pred = "x4",
                   values = c("A", "B"))

calc.oddsratio.gam(model = fit_gam, data = dat, pred = "x4",
                   values = c("B", "D"))
```


## Calculate ORs for percentage increments of predictor distribution (here: 20%)

Slice the training data range in ten pieces and calculate OR between every increment step. 

The first output is the console output giving quick information about the calculated combinations.  

The second output is the `data.frame` that is written if the function is assigned to a an object. 

```{r}
calc.oddsratio.gam(fit_gam, pred = "x2", percentage = 20,
                   slice = TRUE, data = dat, quietly = F)
```

# GLM

```{r, echo=FALSE}
calc.oddsratio.glm <- function(model, data, incr, quietly = FALSE) {
  
  if (class(model)[1] == "glm") {
    
    # get pred names and coefficients without intercept
    preds <- names(coefficients(model))[2:length(coefficients(model))]
    coef <- coefficients(model)[2:length(coefficients(model))]
  }
  
  if (class(model)[1] == "glmmPQL") {
    # get pred names and coefficients without intercept
    preds <- names(model$coefficients$fixed)[2:length(model$coefficients$fixed)]
    coef <- model$coefficients$fixed[2:length(model$coefficients$fixed)]
  }
  
  odds_ratios <- list()
  for (i in preds) {
    # check if predictor is numeric or integer
    if (is.numeric(data[[i]]) | is.integer(data[[i]])) {
      odds_ratios[[i]] <- round(exp(as.numeric(coef[[i]])* as.numeric(incr[[i]])), 3)
      incr1 <- as.numeric(incr[[i]])
      or <- odds_ratios[[i]]
    }
    # if pred is factor -> perform direct conversion to odds ratio
    else {
      odds_ratios[[i]] <- round(exp(as.numeric(coef[[i]])), 3)
      incr1 <- "Non numeric predictor. Refer to basis factor level!"
      or <- odds_ratios[[i]]
    }
    
    if (!quietly) {
      cat("Variable:   '", i, "'\nIncrement:  '", 
          incr1, "'\nOdds ratio: ", or, "\n\n", sep = "")
    }
  }
}
```

# Examples {.tabset}

## GLM

Create example data.   
Data source: http://www.ats.ucla.edu/stat/r/dae/logit.htm

```{r}
dat <- read.csv("http://www.ats.ucla.edu/stat/data/binary.csv")
dat$rank <- factor(dat$rank)
fit_glm <- glm(admit ~ gre + gpa + rank, data = dat, family = "binomial")
```

The function calculates ORs for specific increment steps for all predictors.  
Increments for numerical variabels need to be given in a list.  

Since predictor `rank` is a factor variable with four levels, we will get three odds ratio outputs refering to the basis factor level (here: rank1).  
The output is interpreted as follows: "Having `rank2` instead of `rank1` while holding all other values constant results in a decrease in odds of 49.1% (1-0.509)".

```{r}
calc.oddsratio.glm(fit_glm, dat, list(gre = 380, gpa = 5))
```

## GLMM (MASS::glmmPQL)

Create example data.  
Source: `MASS` package.

```{r}
library(nlme)
library(MASS)
data(bacteria)
fit_glmmPQL <- MASS::glmmPQL(y ~ trt + week, random = ~1 | ID,
                             family = binomial, data = bacteria, verbose = FALSE)
```

```{r}
calc.oddsratio.glm(fit_glmmPQL, bacteria, list(week = 5))
```

